---
title: "Introduction to R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
*QUESTION: How many of you have used R before?* <br>
R is a powerful tool for analyzing and visualizing data
Specifically, it is a language and environment for statistical computing and graphics. As an R user you will need to learn to understand the language R uses to execute commands. And that is what we will hopefully help you with today. 

*QUESTION: How many of you have used RStudio before?* <br>
RStudio is an open source R integrated development environment for interfacing with R.It is useful for:

* Writing scripts for specific analyses
* Creating markdown documents
* Version control via GitHub
* Organizing and sharing analyses via projects

### Navigating RStudio
When we first open RStudio we see that the interface is divided into 3 main windows:

* Console + Terminal
* Environment + History + Connections
* Files + Plots + Packages + Help + Viewer

Once we open a file, such as an R script and editor window will also open on the LHS.

There are two main ways one can work within RStudio.

1. Test and play within the interactive R console then copy code into a `.R` file to run later.

* This works well when doing small tests and initially starting off.
* It quickly becomes laborious 
* Harder to keep track of what you’ve done

In the console window you will see a `>` symbol, which is an interactive command prompt. 

* It operates on the same idea of a “Read, evaluate, print loop”: you type in commands, R tries to execute them, and then returns a result.
* This is similar to what you’d see in a terminal or command prompt. 

2. Start writing in an .R file and use RStudio’s short cut keys for the Run command to push the current line, selected lines or modified lines to the interactive R console.

* This is a great way to start; all your code is saved for later.
* Can easily share code this way.


## Introduction to R

**Using R as a calculator**
Let’s start programming by getting R to do some arithmetic for us. I am going to do this in the editor window. I have linked this file to dropbox, so that you can access all of the coding I am doing via the link in the etherpad. There may be a bit of delay, but I will try to remember to save frequently so that it stays relatively up to date.

There are a few different ways to run commands or 'chunks' of code from the editor window, depending on the OS you are using:

* Click run (all systems)
* Ctrl + Enter (Win, Linux)
* Cmd + Enter (Mac OS)

NOTES: 
* You can run multiple lines of code at once
* You can run lines without highlighting, code will run on line with cursor


So let’s start by running some code to do some addition:

```{r, eval = FALSE}
53 + 23 # [Ctrl + Enter]
```

Remember to actually run this we hit ctrl or cmd + enter, or click the run button.

You’ll see that the line we’ve just run pops up in the console, along with the output from the command we’ve just given R.
If you run an incomplete command, R will wait for you to complete it:

```{r, eval = FALSE}
1 + [Ctrl + Enter]
```

Note that the > in the console window is now a + indicating that are is waiting for the command to be completed. You can exit commands by pressing escape within the console window.

When using R as a calculator, the order of operations is the same as you would have learned back in school. From highest to lowest precedence:

* Parentheses: ( )
* Exponents: ^ or **
* Divide: /
* Multiply: *
* Add: +
* Subtract: -

So for example we can try:
```{r, eval = FALSE}
3 + 5 * 2
```

versus

```{r, eval = FALSE} 
* (3 + 5) * 2
```

*Note that R will ignore spaces between + and - , etc.*


**Mathematical functions**

R has many built in mathematical functions. To call a function, we simply type its name, followed by open and closing parentheses. For example:

```{r, eval = FALSE}
log(1)  # natural logarithm
log10(10) # base-10 logarithm
exp(0.5) # exponent
```

The text after each line of code is called a “comment”. Anything that follows after the hash (or octothorpe) symbol # is ignored by R when it executes code.


**Variables and assignment**

We can store values in variables using the assignment operator`<-`, which is the less than sign followed by a minus like this:

```{r, eval = FALSE}
x <- 1/40
```

*Note: R does not print the value, but x now contains the value 0.025, *
*Note: Will see that x has been loaded in to the environment.*

And we can now perform calculations using x in place of our original number. For example, take the natural log:

```{r, eval=FALSE}
log(x)

```

We can also reassign variables that we have created. Let make x have a value of 100:

```{r, eval=FALSE}
x <- 100
```

We can also update the assigned value by including the variable we want to update. E.g.:

```{r, eval=FALSE}
x <- x + 1
```

and even:

```{r, eval=FALSE}
y <- x * 2
```

Variable names can contain letters, numbers, underscores and periods. They cannot start with a number nor contain spaces at all. Different people use different conventions for long variable names, these include:

* periods.between.words
* underscores_between_words
* camelCaseToSeparateWords

What you use is up to you, but be consistent.

~~~~~~~~~~~~~~~~~~~~~~~~~~
What will be the value of each variable after each statement is the following:

mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age – 20
~~~~~~~~~~~~~~~~~~~~~~~~~~

### Vectorization
R is vectorized: variables and functions can have vectors as values. A vector in R describes a set of values in a certain order of the same data type. For example:

```{r, eval = FALSE}
1:5 
```

Produces a vector of numbers.
And we can perform mathematical functions on this vector:
```{r, eval = FALSE}
2^(1:5)
```

Notice that what R has done here is raise 2 to the power of each number in the vector.

We can also perform arithmetic on multiple vectors, for example
```{r, eval = FALSE}
x <- 1:4
y <-  5:8
x + y
```


## R Packages

It is possible to add functions to R by writing a package, or by obtaining a package written by someone else. Packages are generally written to perform specific functions or statistical analyses, or for manipulating specific types of data. 

R comes with some packages pre-installed, and we can see what packages are installed by typing:

```{r, eval = FALSE}
installed.packages()
```

You can see I have a lot of them and can’t even see the entire list… This will happen when your output is too large to be seen in the console, and is NB to bear in mind when working with large datasets. 

Most packages are available via CRAN: **The Comprehensive R Archive Network**
You install packages by typing `install.packages("packagename")`, where packagename is the package name in quotes. 

These packages are often getting updated, especially as new versions of R come out, and you can update them via `update.packages()`. 

Another key feature is making the library available for use. You can do this through the packages tab, but it is more reproducible to put this into your script. Often people will put this at the top of the script so that you can quickly make sure that you have all of the packages you need to run the code:
To load a package we type:

```{r, eval = FALSE}
library(car)
```

*Note: The library package name only works when you have that package installed, and note that the package name here doesn’t need to be in quotations like it does when you install the package. * <br>
*Note: Once a package is installed, it doesn’t need to be installed again*

Multiple packages can be loaded or installed by concatenating them into a string. For example, you can install `dyplr` and `ggplot2` using:

```{r, eval = F}
install.packages(c("dplyr", "ggplot2"))


```

~~~~~~~~~~~~~~
Install the following packages: 

"ISLR", "ggplot2", "GGally", "dplyr", "cowplot", "mice"

And load them to make them available in the current session.

~~~~~~~~~~~~~~~~~~~~~~~~

## Seeking help
R and any package you load have help files. 
There are a few ways to access help within R and RStudio:

* `?function_name`
* `help(function_name)`


You can also use a fuzzy search by typing `??` for example:

```{r, eval=FALSE}
??read.ta
```

**Special Operators** <br>
You can also get help on special operators, like the assignment operator. Let’s try typing:
```{r, eval = F}
?”<-“
```

Many packages also contain **VIGNETTES**, or tutorials and extended examples. If we type:

```{r, eval = FALSE}
vignette()
```

without any arguments, we will see a list of all vignettes for the installed packages.We can also specify a particular packge:

```{r, eval = FALSE}
vignette(“dplyr”)
```

*Note: we use quotes here*

**When you have no idea where to begin** <br>
If you don’t know what function or package you need to use CRAN Task Views is a list of packages grouped into fields.<br>

Also *Stack Overflow*   can be very helpful. Remeber to use the [r] tag for your question. There are very specific formats for questions on these types of sites; it is suggested that you use the functions:

```{r, eval = FALSE}
?dput
sessionInfo()
```

to help get the information people seeing your question might need.

## Exploring dataframes
